COVID-19 infection and attributable mortality in UK long term care facilities: Cohort study using active surveillance and electronic records (March-June 2020)

### 2020-12-11 manuscript/manuscript.Rmd
resubmission

---
title: "COVID-19 infection and attributable mortality in UK long term care facilities: Cohort study using active surveillance and electronic records (March-June 2020)"
bibliography: references.bib
csl: sage-vancouver.csl
output:
  word_document:
    toc: no
    reference_docx: template.docx
  html_document:
    df_print: paged
    toc: yes
    code_folding: hide
link-citations: yes
always_allow_html: yes
---

Peter Dutey Magni, Researcher$^{12}$ Haydn Williams, Data manager$^{3}$ Arnoupe Jhass, General Practitioner$^{4}$ Greta Rait, Professor$^4$ Fabiana Lorencatto, Senior researcher$^5$ Harry Hemingway,$^{126}$ Andrew Hayward,$^{7}$ Laura Shallcross$^{12*}$

@@

**Background**: Information on the transmission and outcomes of COVID-19 infection in staff and residents in Long term care facilities (LTCFs) in the UK is scarce. This study aimed to estimate the incidence of and risk factors for suspected and confirmed COVID-19 infection and mortality during the first wave of the pandemic.

**Methods and Findings**: 179 UK LTCFs were included in a cohort study using population-level active surveillance and linked electronic health records. Participants included 9,339 residents and 11,604 staff. We estimated the incidence of suspected and confirmed infection and mortality in staff and residents using daily tallies of infection (aggregate data) and used individual-level electronic health records from 8,713 residents to model risk factors for confirmed infection and mortality using Cox proportional hazards. We estimated mortality attributable to COVID-19 by comparing deaths in care homes with and without outbreaks.

`r formatbm(desc_residents$CT_total_symptomatic)`/`r formatbm(desc_residents$CT_total_residents)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory confirmed infections. The rate of incidence of confirmed infection was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days in residents and staff respectively. `r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) LTCFs had at least one COVID-19 infection or COVID-19-related death. Lower staffing ratios and higher occupancy rates were independent risk factors for infection.

Daily tallies from LTCF managers recorded `r desc_residents$DAT_CFR$y` deaths in `r desc_residents$DAT_CFR$n` residents with confirmed infection (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]). Models estimated that 573/1,694 (33.8%) of all-cause deaths were attributable to COVID-19. Mortality in residents with no direct evidence of infection was 98% higher in care homes with at least one COVID-19 infection or death compared to care homes without (adjusted HR 2.0 [1.6; 2.4]).

**Conclusions:** 1 in 5 residents had symptoms of infection during the pandemic, and excess mortality in LTCFs with outbreaks suggest many deaths occurred in people who were infected with COVID-19, but not tested. Higher occupancy and lower staffing levels increase infection risk. Disease control measures should integrate active surveillance and testing with fun-damental changes in staffing and care home occupancy to protect staff and residents from infection.

@@

Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both long-term care facilities (LTCFs) and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 LTCFs for the elderly.[@LaingBuisson2019] Residents of LTCFs are particularly vulnerable to COVID-19 due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and their frequent exposure to infection through close contact with staff members, other residents and contaminated surfaces in the care facility. At the peak of the pandemic, the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020] Staff in LTCFs also have higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] National statistics suggests two-thirds of excess deaths recorded in residents of LTCFs in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested. Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on LTCFs.

Public health strategies to protect the public, residents and staff from COVID-19 include rapid testing, restriction of visitors, and vaccination. Such strategies require knowledge of the burden of and risk factors for infection in residents and staff in LTCFs, linked to outcomes, which draw on evidence from the first wave of the pandemic. Population-based prevalence surveys and studies based on routine data have demonstrated variation in the incidence of infection and case-fatality between countries,[@Fisman2020; @Hollinghurst2020; @Salcher-Konrad2020] but many people with symptoms were not tested, particularly at the start of the pandemic in the context of limited testing capacity. In England, there is no syndromic surveillance for infection in LTCFs in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents in LTCFs until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations which permitted a maximum of five tests per LTCF. Consequently national estimates of incidence and prevalence based on the first wave of infection (February-July 2020) will substantially underestimate the burden of infection in residents and staff in LTCFs.

@@

Electronic record datasets collected by the FSHCG are primarily used for billing and monitoring care quality, but have also been used in previous research studies.[@Smith2020]

@@

FSHCG collects electronic health record data on all their residents except those occupying beds that are 'block contracted' to the local authority (855 beds). The dataset includes: dates of entry to and exit from the LTCF, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, elderly). Individual-level data on incidents including infections are also reported via 'Datix' which records the residents name, LTCF identifier, incident date/time, date of birth, sex, COVID-19 symptoms, tests and test results, resident current location (LTCF/hospital), and death. Individual-level data on residents were linked to Datix reports (supplementary material 1), and used to categorise residents' infection status into four groups:  (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary material 1). The term 'confirmed' denoted a positive PCR test. Datix was also used to differentiate deaths in hospital from those in the LTCF, and to identify COVID-19 related deaths. 1,492/1,880 (79%) of Datix reports were successfully linked.

@@

On 24 March a new reporting system was introduced by FSHCG, which required managers of each LTCF to report daily tallies in residents (new symptomatic cases, new confirmed infection in facility, new confirmed infection in hospital, deaths related to COVID-19) and staff (symptomatic cases, new confirmed cases). The number of occupied beds in each LTCF was reported weekly via the same mechanism. COVID-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to COVID-19 by the coroner. Data on staff deaths were not extracted due to the small number of cases.

Characteristics of each LTCF (number of beds, region, nursing versus residential care) were collected from organisational data. We therefore extracted organisational data, individual-level data for 8,713 residents and aggregate data for all staff and residents (`r fig_nums("fig_study_diagram", display = "cite")`).

@@

## Risk factors for infection

Risk factors for infection included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and LTCF characteristics (number of beds, occupancy, bed to staff ratio, UK-wide Index of Multiple Deprivation [@Abel2016]). Baseline LTCF occupancy was computed by averaging weekly occupancy in January-March 2020, before the first COVID-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms, and modelled as a continuous variable. We also estimated the ratio of beds to staff as a continuous variable. An outbreak in a LTCF was defined as at least one confirmed infection or COVID-19 related death.

@@

Prevalence, incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies. These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] In order to calculate infection and death incidence, we estimated the total number of residents in each care LTCF by extrapolating estimates of LTCF occupancy from the individual-level dataset to the aggregate dataset (because dates of entry and exit to/from the LTCF were not available for 855 beds, supplementary material 1). Daily occupancy was inferred from the weekly report of bed occupancy using linear interpolation. The total number of residents at risk of infection was unknown, so it was approximated in a multiple decrements life table (supplementary material 1). The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and COVID-19 related deaths by day based on the aggregate dataset. The incidence rate ratio for LTCF (based on aggregate data) versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]

Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting (supplementary material 2). The individual-level dataset was used to estimate rates of infection by age and gender, and Cox proportional hazards models were used to test the association between individual and organisational-level risk factors and the outcome of confirmed infection.

@@

Data were analysed in `R`3.5.0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020]

@@

Symptoms of infection were recorded in  `r formatbm(desc_residents$CT_total_symptomatic)` residents based on counts reported by managers of LTCFs, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`). An additional `r desc_residents$CT_total_confirmed` residents had a confirmed infection, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital. The cumulative incidence of confirmed infection was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`]. The rate of confirmed infections in LTCFs in England was 13-fold higher in LTCFs compared to the community prevalence of infection derived from the ONS household COVID-19 infection survey (IRR = `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`]).[@cisbulletin2020]

`r desc_residents$CT_total_cdeath` COVID-19 related resident deaths were reported in the aggregate dataset, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days. `r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of these deaths took place in hospital (`r table_nums("incidence_CT_residents", display = "cite")`).

@@

`r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)` staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) reported symptoms of infection during the study period, and `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).

@@

Estimates of incidence for private residents derived from Datix are reported in supplementary material 2.

@@

Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) residents. `r tbrounding(mean(residents$home_type=="Nursing")*100)`% of residents received nursing care, and 39.2% received dementia care (`r table_nums("tab_resident_descriptives", display = "cite")`). `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in residents of LTCFs, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]. The proportion of resident deaths was two-fold higher in LTCFs with outbreaks compared to those without outbreaks (22.6% versus 11.2%).

@@

Using individual-level resident data, factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model. Male sex, age $\geq$ 85 years, and residence in a nursing LTCF (adjusted HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed COVID-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`). After accounting for organisational differences, LTCF size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for LTCFs with $\geq$ 70 beds versus <35 beds). LTCF baseline occupancy and staffing ratios had the greatest effect on residents' risk of infection. For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR for a 10% point increase = 1.5 [1.1; 2.1]); and a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR for a 10% point increase = 1.3 [1.1; 1.5]).

@@

The time-dependent Cox proportional hazard models in `r table_nums("tab_CoxPH_death", display = "cite")` examine the relationship between infection status (groups A-D) and mortality. After controlling for other risk factors, increased mortality was associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).

We estimated excess mortality in outbreak and non-outbreak LTCFs, taking individuals with no evidence of infection (group A) in non-outbreak LTCFs as the reference group. Hazards of all-cause mortality were almost two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak LTCFs (adjusted HR = 1.9 [1.6; 2.3]). All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.1 [1.9; 5.1]) or symptomatic (Group D: adjusted HR = 12 [9.0; 15]).

@@

Model-based estimates of attributable mortality were derived from the individual-level data. Overall, 573/1,694 (34%) deaths were attributed to COVID-19. In LTCFs with outbreaks only, 160/547 (29%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D, `r table_nums("tab_attributable_deaths", display = "cite")`). Model-based estimates of deaths attributable to COVID-19 based on individual-level data were similar (9% higher) to the number of deaths reported by managers of LTCFs ($n$ = 526) in the aggregate dataset.

@@

This population-level study demonstrates the major impact of COVID-19 on LTCFs, with 22% of residents and 16% of staff experiencing symptoms and a case-fatality rate of 35.7%. Residents with no direct evidence of infection in LTCFs with outbreaks had twice the mortality of the equivalent group in LTCFs without outbreaks, implying substantial case under-ascertainment. Less than one-third of deaths attributable to COVID-19 in outbreak LTCFs were confirmed which is likely due to poor availability of testing until late in the pandemic. In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced LTCF occupancy are critically important to reducing the spread of infection.

>>>>

---
title: "COVID-19 infection and attributable mortality in UK long term care facilities: Cohort study using active surveillance and electronic records (March-June 2020)"
bibliography: references.bib
csl: sage-vancouver.csl
output:
  html_document:
    df_print: paged
    toc: yes
    code_folding: hide
  word_document:
    toc: no
    reference_docx: template.docx
link-citations: yes
always_allow_html: yes
---

Peter Dutey Magni, Researcher$^{1,2}$ Haydn Williams, Data manager$^{3}$ Arnoupe Jhass, General Practitioner$^{4}$ Greta Rait, Professor$^4$ Fabiana Lorencatto, Senior researcher$^5$ Harry Hemingway,$^{1,2,6}$ Andrew Hayward,$^{7}$ Laura Shallcross$^{1,2*}$

@@

**Background**: Epidemiological data on COVID-19 infection in long-term care facilities (LTCFs) are scarce. We analysed data from a large provider of long-term care to investigate infection and mortality during the first wave of the pandemic.

**Methods**: Cohort study of 179 UK LTCFs with  9,339 residents and 11,604 staff. We used manager-reported daily tallies to estimate the incidence of suspected and confirmed infection and mortality in staff and residents.  Individual-level electronic health records from 8,713 residents were used to model risk factors for confirmed infection, mortality, and estimate attributable mortality.

**Results**: `r formatbm(desc_residents$CT_total_symptomatic)`/`r formatbm(desc_residents$CT_total_residents)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory-confirmed infections. The rate of incidence of confirmed infection was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days in residents and staff respectively. `r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) LTCFs had at least one COVID-19 infection or COVID-19-related death. Lower staffing ratios and higher occupancy rates were independent risk factors for infection.

`r desc_residents$DAT_CFR$y` of `r desc_residents$DAT_CFR$n` residents with confirmed infection died (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]). Mortality in residents with no direct evidence of infection was 98% higher in care homes with outbreaks versus those without (adjusted HR 2.0 [1.6; 2.4]).

**Conclusions:** Findings suggest many deaths occurred in people who were infected with COVID-19, but not tested. Higher occupancy and lower staffing levels were independently associated with risks of infection. Protecting staff and residents from infection requires regular testing for COVID-19 and fundamental changes to staffing and LTCF occupancy.

@@

Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both long-term care facilities (LTCFs) and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 LTCFs for the elderly.[@LaingBuisson2019] Residents of LTCFs are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility. At the peak of the pandemic, deaths recorded in LTCF residents were three times higher than during the preceding year.[@onsDR2020] Staff in LTCFs also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] National statistics suggests two-thirds of excess deaths recorded in residents of LTCFs in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested. Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on LTCFs.

Public health strategies to protect the public, residents and staff from COVID-19 include rapid testing, restriction of visitors, and vaccination. Such strategies require knowledge of the burden of and risk factors in residents and staff in LTCFs, linked to outcomes, which may only be drawn from evidence from the pandemic's first wave. Population-based prevalence surveys and studies based on routine data have demonstrated variation in the incidence of infection and case-fatality between countries,[@Fisman2020; @Hollinghurst2020; @Salcher-Konrad2020] but many people with symptoms were not tested, particularly at the start of the pandemic in the context of limited testing capacity. There is no syndromic surveillance for infection in LTCFs in England, and widespread regular testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for staff and residents in LTCFs until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations which permitted a maximum of five tests per LTCF. Consequently national estimates of incidence and prevalence based on the first wave of infection (February–July 2020) substantially underestimate the burden of infection in residents and staff in LTCFs.

@@

We extracted organisational data, individual-level data for 8,713 residents and aggregate data for all staff and residents (`r fig_nums("fig_study_diagram", display = "cite")`). Electronic records collected by the FSHCG are primarily used for billing and monitoring, but have also been used in previous research.[@Smith2020]

@@

FSHCG collects electronic records on all their residents except those occupying beds that are 'block contracted' to the local authority (855 beds). The dataset includes: dates of entry to and exit from the LTCF, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, elderly). Individual-level data on incidents including infections are reported via 'Datix': resident names, LTCF identifier, incident date/time, date of birth, sex, COVID-19 symptoms, test results, resident current location (LTCF/hospital), and death. Individual-level data on residents were linked to Datix reports (supplementary material 1), and used to categorise residents' infection status into four groups:  (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary material 1). The term 'confirmed' denoted a positive PCR test. Datix was also used to differentiate deaths in hospital from those in the LTCF, and to identify COVID-19 related deaths. 1,492/1,880 (79%) of Datix reports were successfully linked.

@@

On 24 March FSHCG introduced a new reporting system requiring managers of each LTCF to report daily tallies in residents (new symptomatic cases, new confirmed infection in facility, new confirmed infection in hospital, deaths related to COVID-19) and staff (new symptomatic cases, new confirmed cases). The number of occupied beds in each LTCF was reported weekly via the same mechanism. COVID-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to COVID-19 by the coroner. Data on staff deaths were not extracted due to the small number of cases.

Characteristics of each LTCF (number of beds, region, nursing versus residential care) were obtained from FSHCG.

@@

## Risk factors

Risk factors considered included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and LTCF characteristics (number of beds, occupancy, bed to staff ratio, UK-wide Index of Multiple Deprivation [@Abel2016]). Baseline LTCF occupancy was computed by averaging weekly occupancy in January-March 2020, before the first COVID-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms, and modelled as a continuous variable. We also estimated the ratio of beds to staff as a continuous variable. An outbreak in a LTCF was defined as at least one confirmed infection or COVID-19 related death.

@@

Prevalence, incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies. These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff (`r fig_nums("fig_study_diagram", display = "cite")`).[@CQC2020stats] Daily occupancy and numbers of residents at risk of infection were inferred using interpolation and a life table approach, as described in supplementary material 1. Daily occupancy was inferred from the weekly report of bed occupancy using linear interpolation. The total number of residents at risk of infection was unknown, so it was approximated in a multiple-decrement life table (supplementary material 1). The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and COVID-19 related deaths by day based on the aggregate dataset. The incidence rate ratio for LTCF (based on aggregate data) versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]

Infection incidence was also estimated from the individual-level dataset, but was subject to under-reporting (supplementary material 2). Due to this, individual-level data were only used to estimate age/sex-specific rates of infection and Cox proportional hazards models testing the association with individual and organisational-level risk factors.

@@

Data were analysed in `R`3.5.0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. Computer scripts are available online.[@Dutey2020]

@@

LTCF managers recorded symptoms of infection in  `r formatbm(desc_residents$CT_total_symptomatic)` residents, contributing to an overall cumulative incidence of `r tbrounding(desc_residents$CT_p_symptomatic)`% [`r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%] or an incidence rate of `r tbrounding(desc_residents$CT_IR_symptomatic)` per 100,000 resident-days [`r tbrounding(desc_residents$CT_IR_symptomatic_lower)`; `r tbrounding(desc_residents$CT_IR_symptomatic_upper)`] (`r table_nums("incidence_CT_residents", display = "cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`). An additional `r desc_residents$CT_total_confirmed` residents had a confirmed infection, of whom `r formatbm(sum(new_cases$new_confirmed_hospital, na.rm = T))` were diagnosed in hospital. The cumulative incidence of confirmed infection was `r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%], with an incidence rate of `r tbrounding(desc_residents$CT_IR_confirmed)` per 100,000 [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`]. The rate of confirmed infections in LTCFs in England was 13-fold higher in LTCFs compared to the community prevalence of infection derived from the ONS household COVID-19 infection survey (IRR = `r tbrounding(england_results$results$measure[2,"estimate"])` [`r tbrounding(england_results$results$measure[2,"lower"]) `; `r tbrounding(england_results$results$measure[2,"upper"])`]).[@cisbulletin2020]

LTCF managers recorded `r desc_residents$CT_total_cdeath` COVID-19 related resident deaths, equivalent to a crude incidence of `r tbrounding(desc_residents$CT_p_cdeath)`% [`r tbrounding(desc_residents$CT_p_cdeath_lower)`; `r tbrounding(desc_residents$CT_p_cdeath_upper)`] or `r  tbrounding(desc_residents$CT_IR_cdeath)` [`r  tbrounding(desc_residents$CT_IR_cdeath_lower)`; `r  tbrounding(desc_residents$CT_IR_cdeath_upper)`] per 100,000 resident-days. `r tbrounding(sum(new_cases$new_deaths_hospital, na.rm = T)/desc_residents$CT_total_cdeath*100)`% of these deaths took place in hospital (`r table_nums("incidence_CT_residents", display = "cite")`).

@@

LTCF managers recorded `r formatbm(desc_staff$CT_symptomatic_total)`/`r formatbm(desc_staff$total_staff)` staff (`r tbrounding(desc_staff$CT_symptomatic_p)`% [`r tbrounding(desc_staff$CT_symptomatic_p_lower)`%; `r tbrounding(desc_staff$CT_symptomatic_p_upper)`%]) experiencing symptoms of infection during the study period, while `r formatbm(desc_staff$CT_confirmed_total)` (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had a confirmed infection (`r table_nums("incidence_CT_staff", display="cite")`, `r fig_nums("fig_KM_CT_plots", display = "cite")`).

@@

Incidence estimates derived from individual-level Datix records are reported in supplementary material 2.

@@

Individual-level data were available for `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_total_residents/desc_residents$CT_total_residents*100)`%) private residents (`r table_nums("tab_resident_descriptives", display = "cite")`), who accounted for `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths, equivalent to a crude cumulative incidence of `r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]. The proportion of resident deaths was two-fold higher in LTCFs with outbreaks compared to those without outbreaks (22.6% versus 11.2%).

@@

Factors affecting the rate of confirmed cases were investigated in a Cox Proportional Hazard model. Male sex, age $\geq$ 85 years, and residence in a nursing LTCF (adjusted HR = 1.6 [1.0; 2.4]) were all independently associated with increased risk of confirmed COVID-19 infection (`r table_nums("tab_CoxPH_confirmed", display = "cite")`). After controlling for organisational differences, LTCF size no longer had a statistically significant association with rates of infection (adjusted HR = 1.7 [0.7; 4.3] for LTCFs with $\geq$ 70 beds versus <35 beds). LTCF baseline occupancy and staffing ratios had the greatest effect on residents' risk of infection. For example, a 10 percentage point increase in the ratio of occupants to bedrooms was associated with a 51% increase in infection (adjusted HR = 1.5 [1.1; 2.1]); and a 10 percentage point increase in the ratio of beds to staff was associated with a 26% increase in infection (adjusted HR = 1.3 [1.1; 1.5]).

@@

Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality. After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).

We estimated excess mortality in outbreak and non-outbreak LTCFs, taking individuals with no evidence of infection (group A) in non-outbreak LTCFs as the reference group. Hazards of all-cause mortality were almost two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak LTCFs (adjusted HR = 1.9 [1.6; 2.3]). All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.1 [1.9; 5.1]) or symptomatic (Group D: adjusted HR = 12 [9.0; 15]). In confirmed infections, mortality was significantly higher in individuals a record of symptoms.

@@

Model-based estimates indicate that 573/1,694 (34%) all-cause deaths were attributable to COVID-19 (`r table_nums("tab_attributable_deaths", display = "cite")`). In LTCFs with outbreaks only, 160/547 (29%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D, `r table_nums("tab_attributable_deaths", display = "cite")`).

@@

Across 179 LTCFs, 22% of residents experienced symptoms while 10% had a laboratory-confirmed infection, with a case-fatality rate of 35.7% across the first wave of the pandemic. Residents with no direct evidence of infection in LTCFs with outbreaks had twice the mortality of the equivalent group in LTCFs without outbreaks: less than one-third of deaths attributable to COVID-19 in outbreak LTCFs were confirmed due to insufficient testing capacity until late in the pandemic. In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced LTCF occupancy are critically important to reducing the spread of infection.

Our estimates of the prevalence of confirmed COVID-19 infections and deaths in residents are comparable to a large survey of managers of LTCFs in England.[@onsVivaldi2020] However, both studies are likely to be underestimates due to limited testing, asymptomatic infection[@Salcher-Konrad2020] and moderate sensitivity of PCR testing.[@Kucirka2020] Our estimate of `r tbrounding(desc_residents$DAT_CFR_rate)`% case-fatality in residents with confirmed infection over a mean of 71 days is slightly higher than previous literature,[@Blain2020; @McMichael2020; @Arons2020] but is based on longer follow-up, a larger number of residents, and our study population had higher overall mortality.

Two thirds of LTCFs in our study reported at least one case of infection or death, in agreement with a recent study from the Lothian region of Scotland which reported that 61% of LTCFs for older adults had experienced an outbreak.[@Burton2020] This suggests that most outbreaks were identified through FSHCs active surveillance system, and supports our assumption that residents in non-outbreak LTCFs had not been exposed to infection. This assumption made it possible to estimate mortality attributable to COVID-19.
